---
title: "Analyzing Inflation Dynamics in Canada During and After the COVID-19 Pandemic: A Deep Dive into Consumer Price Index Movements from 2019 to 2023"
format: pdf
author: 
- Qin He
abstract: "The COVID-19 pandemic has left a significant imprint on global economic structures, influencing inflation rates worldwide. Canada, like many countries, experienced fluctuating inflation rates during and post-pandemic, marking a period of economic instability and adjustment. This paper delves into the inflation dynamics in Canada from 2019 to 2023, with a special focus on the consumer price index (CPI) as a reflection of the cost of goods and services impacting the everyday lives of Canadians. It explores the trajectory of the overall inflation rate, which peaked in 2023, highlighting the role of various sectors, including food, housing, and energy, in driving these trends. By analyzing data from Statistics Canada, this study provides insights into how the pandemic and subsequent recovery phases have affected inflation. It examines the interplay between supply chain disruptions, monetary policy responses, and changes in consumer behavior, culminating in a comprehensive overview of inflation's impact on the Canadian economy. This analysis not only sheds light on the inflationary trends but also contributes to understanding the broader economic implications of the COVID-19 pandemic on Canada."
date: today
date-format: long
editor: visual
toc: TRUE
toc-depth: 1
number-sections: TRUE
bibliography: references.bib
thanks: "See the GitHub repository supporting this analysis: https://github.com/qinheinfo/Canada_inflation"
---

```{r}
#| include: false
#| warning: false
#| message: false
#| echo: false
#| output: false

library(tidyverse)
library(palmerpenguins)
library(lubridate)
library(dplyr)
library(knitr)
library(janitor)
library(scales)
library(RColorBrewer)
library(ggplot2)
library(kableExtra)
library(here)
library(arrow)

# Fetch the the clean inflation data
inflation_data <- arrow::read_parquet(here::here("----"), show_col_types = FALSE)

# Fetch clean CPI grocery item data
grocery_data <- arrow::read_parquet(here::here("----"), show_col_types = FALSE)
```

```


# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....



# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it. 



# Data

## Measurement
xxx

## Data Analysis


# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```




# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows... 

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```



\newpage


# References


